### 2024-06-15 research-methods.qmd
20240615 - mediation/moderation figures
20240615 - typo
20240615 - correlation and causation

@@

A *longitudinal study* is a study with more than one timepoint.
When the <u>same measures<u> are assessed at each of multiple timepoints, we refer to this as a "repeated measures" design.

@@

[Correlational designs](#sec-correlationalStudy) have greater capacity for [external validity](#sec-externalValidity) than [experimental designs](#sec-experiment) because the participants can be observed in their natural environments to evaluate how variables are related in the real world.
However, the greater [external validity](#sec-externalValidity) comes at a cost of lower [internal validity](#sec-internalValidity).
[Correlational designs](#sec-correlationalStudy) are not well-positioned to make causal inferences.
As the maxim goes, "correlation does not mean causation"—just because two variables are associated does not necessarily mean that they are causally related.
There could be third variables (i.e., confounds) that act as a common cause of both the predictor and outcome variables, thus explaining why they are correlated.

For instance, it may not be that sport drink consumption per se influences player performance; rather, it may be that players who are more intelligent or have more financial resources tend to drink more sports drinks and also tend to perform better.
In this case, intelligence or financial resources may be a confound that influences both sports drink consumption and player performance, but sports drink consumptions—though correlated with player performance—does not influence player performance.
Or, the variables could be associated due to the opposite direction of effect (i.e., the outcome variable may influence the predictor variable).
For instance, greater performance may lead players to drink more sports drink (rather than the reverse).
[Correlational studies](#sec-correlationalStudy) can account for potential confounds using *covariates* or for the reverse direction of effect using longitudinal designs, but the researcher has less control over the variables than in an [experiment](#sec-experiment).

@@

![Mediation Versus Moderation Mnemonic](images/mediationModerationMnemonic.png)

@@

**Mediation** is a <u>causal chain of events</u>, where one variable (a mediator variable) at least partially <u>explains (or accounts for) the association between two other variables</u> (the predictor variable and the outcome variable).
In mediation, a predictor (`X`) leads to a mediator (`M`), which leads to an outcome (`Y`).
Mediation answers the question of, "<u>**Why (or how)** does `X` influence `Y`?</u> A <u>mediator</u> (`M`) is a variable that helps explain the assocation between two other variables, and it answers the question of why/how `X` influences `Y`.
That is, the mediator is the variable that helps explain how/why `X` is related to `Y`.
In other words, you can think of the mediator as the <u>mechanism</u> that helps explain why `X` has an impact on `Y`.
The association between `X` and `Y` <u>gets smaller</u> when accounting for `M`.
Visually this can be written as:

![Mediation](images/mediation.png)

@@

As a picture, we can draw this assocation as:

[INSERT]

@@

When one mechanism fully accounts for the effect of the predictor variable on the outcome variable, this is known as **full mediation**, as depicted below:

![Full Mediation](images/mediation.png)

@@

When a single process partially—but does not fully—accounts for the effect of the predictor variable on the outcome variable; this is known as **partial mediation** and is depicted below:

![Partial Mediation](images/mediationPartial.png)

@@

In addition, there can be multiple mediators/mechanisms that account for the effect of a predictor variable on an outcome variable, as depicted below:

![Multiple Mediators](images/mediationMultiple.png)

@@

**Moderation** (sometimes called an "interaction"), on the other hand, occurs when there is a variable or condition (`M`; called a "moderator") that <u>changes the assocation</u> between `X` and `Y`.
That is, the effect of the predictor variable on the outcome variable differs at different levels of the moderator variable.
In these cases, `X` and `M` <u>work together</u> to have an effect on `Y`; here `X` <u>does not have a direct effect</u> on `M`.
Moderation answers the question of, "<u>**For whom** does `X` influence `Y`?</u>"
If `X` influences `Y` more strongly for some people or in some circumstances, we would say that there is an interaction such that the <u>effect of `X` on `Y` depends on `M`</u>, as depicted below:

![Moderation](images/moderation.png)

For example, if the effect of consuming sports drinks on player performance differs for Quarterbacks and Wide Receivers, the interaction could be depicted below:

[INSERT]

>>>>

#### Correlation Does Not Imply Causation {#sec-correlationCausation}

As the maxim goes, "correlation does not imply causation"—just because two variables are associated does not necessarily mean that they are causally related.

Just because $X$ is associated with $Y$ does not mean that $X$ causes $Y$.
Consider that you find an association between variables $X$ and $Y$.

There are several reasons why you might observe an association between $X$ and $Y$:

- $X$ causes $Y$
- $Y$ causes $X$
- a third variable (i.e., confound), $Z$, influences both $X$ and $Y$
- the association between $X$ and $Y$ is spurious

For instance, one possibility is that the association we observed reflects our hypothesis that $X$ causes $Y$, as depicted in @fig-correlationAndCausation1.
That is, consumption of more sports drink may improve players' performance.

::: {#fig-correlationAndCausation1}
![](images/correlationAndCausation_1.png)

Hypothesized Causal Effect Based on an Observed Association Between $X$ and $Y$, Such That $X$ Causes $Y$.
:::

However, a second possibility is that the association reflects the opposite direction of effect, where $Y$ actually causes $X$, as depicted in @fig-correlationAndCausation2.
For instance, greater performance may lead players to drink more sports drink (rather than the reverse).

::: {#fig-correlationAndCausation2}
![](images/correlationAndCausation_2.png)

Reverse (Opposite) Direction of Effect From the Hypothesized Effect, Where $Y$ Causes $X$.
:::

A third possibility is that the association could reflect the influence of a third variable.
If a third variable is a common cause of each and accounts for their association, it is a *confound*.
An observed association between $X$ and $Y$ could reflect a confound—i.e., a cause ($Z$) that influences both $X$ and $Y$, which explains why $X$ and $Y$ are correlated even though they are not causally related.
A third variable confound that is a common cause of both $X$ and $Y$ is depicted in @fig-correlationAndCausation3.
For instance, it may not be that sport drink consumption per se influences player performance; rather, it may be that players who are more intelligent or have more financial resources tend to drink more sports drinks and also tend to perform better.
In this case, intelligence or financial resources may be a confound that influences both sports drink consumption and player performance, but sports drink consumptions—though correlated with player performance—does not influence player performance.

::: {#fig-correlationAndCausation3}
![](images/correlationAndCausation_3.png)

Confounded Association Between $X$ and $Y$ due to a Common Cause, $Z$.
:::

Lastly, the association might be spurious.
It might just reflect random variation (i.e., chance), and that when tested on an independent sample, what appeared as an association in the original dataset may not hold when testing the association in a new dataset.

@@

A *longitudinal study* is a study with more than one timepoint.
When the <u>same measures</u> are assessed at each of multiple timepoints, we refer to this as a "repeated measures" design.

@@

[Correlational designs](#sec-correlationalStudy) have greater capacity for [external validity](#sec-externalValidity) than [experimental designs](#sec-experiment) because the participants can be observed in their natural environments to evaluate how variables are related in the real world.
However, the greater [external validity](#sec-externalValidity) comes at a cost of lower [internal validity](#sec-internalValidity).
[Correlational designs](#sec-correlationalStudy) are not well-positioned to make causal inferences.
[Correlational studies](#sec-correlationalStudy) can account for potential confounds using *covariates* or for the reverse direction of effect using longitudinal designs, but the researcher has less control over the variables than in an [experiment](#sec-experiment).

@@

A mnemonic to help remember the difference between [mediation](#mediation) and [moderation](#moderation) is in @fig-mediationModerationMnemonic.

::: {#fig-mediationModerationMnemonic}
![](images/mediationModerationMnemonic.png)

Mediation Versus Moderation Mnemonic.
:::

@@

**Mediation** is a <u>causal chain of events</u>, where one variable (a mediator variable) at least partially <u>explains (or accounts for) the association between two other variables</u> (the predictor variable and the outcome variable).
In mediation, a predictor (`X`) leads to a mediator (`M`), which leads to an outcome (`Y`).
Mediation answers the question of, "<u>**Why (or how)** does `X` influence `Y`?</u> A <u>mediator</u> (`M`) is a variable that helps explain the assocation between two other variables, and it answers the question of why/how `X` influences `Y`.
That is, the mediator is the variable that helps explain how/why `X` is related to `Y`.
In other words, you can think of the mediator as the <u>mechanism</u> that helps explain why `X` has an impact on `Y`.
The association between `X` and `Y` <u>gets smaller</u> when accounting for `M`.
Visually this can be written as in @#fig-mediation:

::: {#fig-mediation}
![](images/mediation.png)

Mediation.
:::

@@

As a picture, we can draw this assocation as in @fig-mediationExample:

::: {#fig-mediationExample}
![](images/mediationExample.png)

Mediation Example.
:::

@@

When one mechanism fully accounts for the effect of the predictor variable on the outcome variable, this is known as **full mediation**, as depicted in @fullMediation:

::: {#fig-fullMediation}
![](images/mediation.png)

Full Mediation.
:::

@@

When a single process partially—but does not fully—accounts for the effect of the predictor variable on the outcome variable; this is known as **partial mediation** and is depicted in @fig-fullMediation:

::: {#fig-fullMediation}
![](images/mediationPartial.png)

Partial Mediation.
:::

@@

In addition, there can be multiple mediators/mechanisms that account for the effect of a predictor variable on an outcome variable, as depicted in @fig-multipleMediators:

::: {#fig-multipleMediators}
![](images/mediationMultiple.png)

Multiple Mediators.
:::

@@

**Moderation** (sometimes called an "interaction"), on the other hand, occurs when there is a variable or condition (`M`; called a "moderator") that <u>changes the assocation</u> between `X` and `Y`.
That is, the effect of the predictor variable on the outcome variable differs at different levels of the moderator variable.
In these cases, `X` and `M` <u>work together</u> to have an effect on `Y`; here `X` <u>does not have a direct effect</u> on `M`.
Moderation answers the question of, "<u>**For whom** does `X` influence `Y`?</u>"
If `X` influences `Y` more strongly for some people or in some circumstances, we would say that there is an interaction such that the <u>effect of `X` on `Y` depends on `M`</u>, as depicted in @fig-moderation:

::: {#fig-moderation}
![](images/moderation.png)

Moderation.
:::

For example, if the effect of consuming sports drinks on player performance differs for Quarterbacks and Wide Receivers, the interaction could be depicted in @moderationExamplePathDiagram and @fig-moderationExampleInteractionGraph:

::: {#fig-moderationExamplePathDiagram}
![](images/moderationExample1.png)

Moderation Example: Path Diagram.
:::

::: {#fig-moderationExampleInteractionGraph}
![](images/moderationExample2.png)

Moderation Example: Interaction Graph.
:::
